---
output: github_document
title: EJAM Environmental Justice Analysis Multisite tool
---

<!-- README.md is generated from README.Rmd. Please edit Rmd not md  -->

```{r, echo=FALSE, eval=FALSE}
# Developer note: 
# If you use index.Rmd or README.Rmd it is your responsibility to 
# knit the document to create the corresponding .md. 
# pkgdown does not do this for you because it only touches 
# files in the ⁠doc/⁠ directory.
#
# 1) Edit your README.Rmd file (the .Rmd NOT the .md file)
# 2) Build your README.md file from the README.Rmd file, by running 
# 
#
#        devtools::build_readme()
#
#
# 3) Commit both your README.Rmd and README.md
#
# NOTE: developers need to render `README.Rmd` regularly, to keep `README.md` up-to-date. You could also use GitHub Actions to re-render
# `README.Rmd` every time you push. An example workflow can be found here:
# <https://github.com/r-lib/actions/tree/v1/examples>.
#
# Then... 
#    - **See ?pkgdown::build_site** - EJAM now uses the pkgdown R package to build help and vignettes as web pages, and it uses the README.md (not .Rmd) to help build the documentation.
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

<!-- badges: end -->

[The EJAM Web App] makes it easy to see demographic and environmental information summarized in and across any list of places in the nation. Using EJAM is like running an EJScreen report, but for hundreds or thousands of places, all at the same time. The web app lets you quickly check a summary report, explore interactive maps, tables, and plots, and download a summary or detailed spreadsheet.

[The EJAM R Package for Analysts and Developers] written in the R programming language, with source code on GitHub. It provides functions to help work with block group data, points, and polygons, to very quickly aggregate and compare large numbers of locations.

EJAM is provided by the US Environmental Protection Agency (US EPA).

## The EJAM Web App

-   [What is EJAM?](articles/0_what_is_ejam.html "0_what_is_ejam.html")

-   [Using EJAM as a Web App](articles/0_webapp.html "0_webapp.html")

-   [EJAM live beta test](https://rstudio-connect.dmap-stage.aws.epa.gov/content/cb2bc9cf-75c3-4db1-ad28-e92940fe4438/ "https://rstudio-connect.dmap-stage.aws.epa.gov/content/cb2bc9cf-75c3-4db1-ad28-e92940fe4438/")\*\* (must be at EPA offices or on VPN)

## The EJAM R Package for Analysts and Developers

-   [1. Installing the EJAM R package](articles/1_installing.html "1_installing.html")

-   [2. Quick Start Guide to the EJAM R package](articles/2_quickstart.html "2_quickstart.html")

-   [3. Basics of Using EJAM for Analysis in RStudio](articles/3_analyzing.html "3_analyzing.html")

-   [4. Advanced Features in the EJAM R package](articles/4_advanced.html "4_advanced.html")

-   [5. Accessing a Batch of EJScreen Reports via EJAM](articles/5_ejscreenapi.html "5_ejscreenapi.html")

-   [6. Future Plans for EJAM](articles/6_future_plans.html "6_future_plans.html")

-   [Reference Manual Documenting all EJAM Package Functions and Data](reference/index.html "index.html")

-   [GitHub Repository - Code and Data, at USEPA/EJAM](https://github.com/USEPA/EJAM#readme "https://github.com/USEPA/EJAM#readme") (via EPA VPN or LAN in-office)

## Disclaimer for App

This software/application has been approved for release by the U.S. Environmental Protection Agency (USEPA). Although the software has been subjected to rigorous review, the USEPA reserves the right to update the software as needed pursuant to further analysis and review. No warranty, expressed or implied, is made by the USEPA or the U.S. Government as to the functionality of the software and related material nor shall the fact of release constitute any such warranty. Furthermore, the software is released on condition that neither the USEPA nor the U.S. Government shall be held liable for any damages resulting from its authorized or unauthorized use.

## Disclaimer for GitHub Repository Content

The United States Environmental Protection Agency (EPA) GitHub project code is provided on an "as is" basis and the user assumes responsibility for its use. EPA has relinquished control of the information and no longer has responsibility to protect the integrity, confidentiality, or availability of the information. Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by EPA. The EPA seal and logo shall not be used in any manner to imply endorsement of any commercial product or activity by EPA or the United States Government.
