---
title: "Using EJAM"
description: "Using EJAM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using EJAM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r developernote, eval=FALSE, echo= FALSE, include = FALSE}
#  *>>>>>>>>>> Developer note: vignettes need to be tested/edited/rebuilt regularly <<<<<<<<<<<*
#    - **See ?pkgdown::build_site** and script in EJAM/data-raw/- EJAM uses the pkgdown R package to build help and articles/ vignettes as web pages
```

```{r SETUP_default_eval_or_not, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(eval = FALSE)
# https://r-pkgs.org/vignettes.html
```

# Web app for the public via EJScreen

*EPA is working to integrate EJAM's new toolkit into EJScreen*, the public web app, to make public these features as soon as possible. EPA's [EJScreen](https://www.epa.gov/ejscreen "https://www.epa.gov/ejscreen"){.uri target="_blank" rel="noreferrer noopener"} will leverage EJAM's toolkit to provide public access to batches of EJ reports, with rapid multi-site analysis.

# Web app for EPA staff

EPA staff can access an [internal version of the EJAM web app](https://usepa.sharepoint.com/sites/EJAMWorkgroup/SitePages/access-ejam.aspx). It is for use by staff at EPA offices or via VPN, and requires authentication with a PIV card.

```{=html}
<!--
As of October 2024, EPA staff can use version 2.32 of the EJAM web app - The tool is accessible only at EPA offices or via VPN -- You must be inside EPA's network (e.g. using VPN) to access the EJAM web app, and need to authenticate with PIV card.
-->
```
# Web app for analysts or developers using R/RStudio

*Once the EJAM software and data are made available as open source resources*, anyone (not just EPA staff) will be able to use EJAM in RStudio on their own computer.

EPA staff (on-site or on the VPN) can access the internal version of the EJAM repository as explained [here](https://usepa.sharepoint.com/sites/EJAMWorkgroup/SitePages/access-ejam.aspx).

*Assuming you have access to the code repository, the following will apply:*

Analysts or developers using R/RStudio have the option of running a local copy of the EJAM web app on their own computer. This may be even faster than relying on the hosted web app, does not time out after inactivity, and could be customized by a developer.

If you use the R programming language in RStudio, you can use EJAM through a web-based UI run on your own computer (and/or use EJAM functions and data directly without the web app, for more complex work).

You first need to install the EJAM R package and datasets as explained in [Installing the EJAM R package](https://usepa.github.io/EJAM/articles/1_installing.html). There is also a [Quick Start guide](https://usepa.github.io/EJAM/articles/2_quickstart.html) and extensive [documentation of EJAM functions/tools/data](https://usepa.github.io/EJAM/reference/index.html).

Once EJAM is installed (and you have access to the datasets needed), you can launch the local web app from RStudio as follows:

```{r run_app1, eval = FALSE, include= TRUE}
library(EJAM)
run_app()
```

Note this is different than running a simple shiny app that is not also a package via the golem package. You should use the EJAM package function `run_app()` not `shiny::runApp()`.

Note that the web app (e.g., launched locally) handles loading datasets and indexing blocks, assuming you have access to the server that stores those datasets as a pins board, so you would not have to download datasets manually or do indexing manually if launching the EJAM app locally. You can also download the datasets and store them locally. You do need the packages installed that EJAM relies on, as listed in the DESCRIPTION file.

```{=html}
<!--
EJAM has a series of high-level functions that an analyst can use in RStudio - EJAM can be installed and then used in RStudio as explained in the articles on Installing the EJAM R package and the Quick Start Guide.
-->
```
