---
title: "0. Using EJAM as a Web App"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{0. Using EJAM as a Web App}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r SETUP_default_eval_or_not, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(eval = FALSE)
# https://r-pkgs.org/vignettes.html
```

```{r developernote, eval=FALSE, echo= FALSE, include = FALSE}
#
#  *>>>>>>>>>> Developer note: vignettes need to be tested/edited/rebuilt regularly <<<<<<<<<<<*
#
#    - **See ?pkgdown::build_site** - EJAM now uses the pkgdown R package to build help and vignettes as web pages
# 
#     - It previously used some of these methods also:
#       - rmarkdown::render("vignettes/INSERTNAMEHERE.Rmd", encoding = 'UTF-8')
#       - The "render button" in RStudio that shows more detailed progress during render
#       - devtools::build_vignettes(), but that failed on dataload_from_pins() or _local() trying to get blockid2fips
```

# Using the EJAM web app

As of February 2024, EPA staff can try a live beta test version of EJAM. You must be inside EPA's network (e.g. using VPN) to access the EJAM web app, currently. EJAM is still in active development, and is a work in progress, so not everything is bug-free yet and changes are being made, so we are seeking feedback and input on bugs, usability issues, performance, clarity, and ideas for enhancements with possible new features and functionality.

-   The first complete version of EJAM is being beta tested on a staging server, accessible only by EPA staff members using VPN/in-LAN and after signing in. [**EJAM beta test**](https://rstudio-connect.dmap-stage.aws.epa.gov/content/cb2bc9cf-75c3-4db1-ad28-e92940fe4438)

Information about how to use the web app may be provided elsewhere at some point, but the app is meant to be relatively easy for an analyst to use without any training or user guide.

# Running the EJAM web app in R/RStudio on your own computer (for analysts or developers)

If you use R/RStudio, you can run the web app on your own computer. You first need to install the EJAM R package as explained in the article/vignette on installing EJAM.

Once EJAM is installed (and you have access to the datasets needed), you can launch the local web app from RStudio as follows:

```{r run_app1, eval = FALSE, include= TRUE}
library(EJAM)
run_app()
```

Note this is different than running a simple shiny app that is not also a package via golem. You should use the EJAM package function `run_app()` not `shiny::runApp()`.

Note that the web app (e.g., launched locally) handles loading datasets and indexing blocks, assuming you have access to the server that stores those datasets as a pins board, so you would not have to download datasets manually or do indexing manually if launching the EJAM app locally. You just need the packages installed that EJAM relies on (see below).

# Using EJAM tools in R/RStudio, outside the web app

For those wanting to go beyond the web app, to use EJAM functions directly in R/RStudio, see the [Quick Start Guide](docs/articles/2_quickstart.html "docs/articles/2_quickstart.html").
