---
title: "Using EJAM"
description: "Using EJAM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using EJAM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r developernote, eval=FALSE, echo= FALSE, include = FALSE}
#  *>>>>>>>>>> Developer note: vignettes need to be tested/edited/rebuilt regularly <<<<<<<<<<<*
#    - **See ?pkgdown::build_site** and script in EJAM/data-raw/- EJAM uses the pkgdown R package to build help and articles/ vignettes as web pages
```

```{r SETUP_default_eval_or_not, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(eval = FALSE)
# https://r-pkgs.org/vignettes.html
```

# EJAM and EJScreen websites

EJAM itself is not an app running on a public website. [EJScreen](https://www.epa.gov/ejscreen "https://www.epa.gov/ejscreen"){.uri target="_blank" rel="noreferrer noopener"} will be able to leverage EJAM's technology to provide public access to batches of reports, and will be able to do rapid multi-site analysis in a future update. EPA is working to integrate EJAM's new toolkit into EJScreen's public screening and mapping web application, as a way to enable these features as soon as possible. A web app was being tested for internal use at EPA as of early 2024.

<!--
As of mid 2024, EPA staff can try a live beta test version of the EJAM web app - The beta test is accessible only at EPA offices or via VPN.

You must be inside EPA's network (e.g. using VPN) to access the EJAM web app, and need to authenticate with PIV card, currently.

EJAM is still in active development, and is a work in progress, so not everything is bug-free yet and changes are being made, so we are seeking feedback and input on bugs, usability issues, performance, clarity, and ideas for enhancements with possible new features and functionality.

Information about how to use the web app may be provided elsewhere at some point, but the app is meant to be relatively easy for an analyst to use without any training or user guide.
-->
# R/RStudio users launching a web app

If you use R/RStudio, you also can use EJAM through a web-based UI on your own computer. You first need to install the EJAM R package and datasets as explained in Installing the EJAM R package.

Once EJAM is installed (and you have access to the datasets needed), you can launch the local web app from RStudio as follows:

```{r run_app1, eval = FALSE, include= TRUE}
library(EJAM)
run_app()
```

Note this is different than running a simple shiny app that is not also a package via the golem package. You should use the EJAM package function `run_app()` not `shiny::runApp()`.

Note that the web app (e.g., launched locally) handles loading datasets and indexing blocks, assuming you have access to the server that stores those datasets as a pins board, so you would not have to download datasets manually or do indexing manually if launching the EJAM app locally. You can also download the datasets and store them locally. You do need the packages installed that EJAM relies on, as listed in the DESCRIPTION file, including the ones not on CRAN, notably EJAMejscreenapi and EJAMbatch.summarizer.

<!--
EJAM has a series of high-level functions that an analyst can use in RStudio - EJAM can be installed and then used in RStudio as explained in the articles on Installing the EJAM R package and the Quick Start Guide.
-->
