---
title: "What is EJAM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What is EJAM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r developernote, eval=FALSE, echo= FALSE, include = FALSE}
#  *>>>>>>>>>> Developer note: vignettes need to be tested/edited/rebuilt regularly <<<<<<<<<<<*
#    - **See ?pkgdown::build_site** - EJAM uses the pkgdown R package to build help and articles/ vignettes as web pages
```

```{r SETUP_default_eval_or_not, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(eval = FALSE)
# https://r-pkgs.org/vignettes.html
```

## EPA's EJ Analysis Multi-site (EJAM) tool

EJAM is a tool, developed by the
United States Environmental Protection Agency (US EPA)
that makes it easy to see demographic and environmental information 
summarized in and across any list of places in the nation. 
Using EJAM is like running an EJScreen report, 
but for hundreds or thousands of places, 
all at the same time. 

EJScreen will be able to leverage EJAM's technology to 
let you quickly check a summary report, 
explore interactive maps, tables, and plots, and 
download a summary or detailed spreadsheet.

This provides interactive results and a formatted, ready-to-share report with written explanations of the results, tables, and graphics. 
The report can provide EJ-related information about people who live in communities near any of the industrial facilities on a list, for example.

## Overview

EJAM is a user-friendly web application that can summarize demographics and environmental conditions for any list of places in the nation.
It provides interactive results and a formatted, ready-to-share report with written explanations of the results, tables, and graphics.
The report can provide EJ-related information about people who live in communities near any of the industrial facilities on a list, for example.

To use the tool, one first selects the places to be analyzed.
This could be, for example, everyone within 2 miles of any EPA-regulated facility in a specific NAICS code (industrial sector).
The tool runs a fast "buffer" or proximity analysis at each location, 
similar to how EJScreen provides a standard report for a single location, 
except EJAM does this for each of a large number of locations very quickly. 
Then, most importantly, EJAM provides a complete overview report, 
to summarize environmental conditions and demographics 
across all the populations and all of the locations. 
The results can be explored interactively or downloaded as a written report with text, tables, and graphics.

An EJAM report can quickly and easily show which demographic groups live near the selected facilities. 
It also provides new insights into which environmental stressors 
may affect certain demographic subgroups disproportionately, to varying degrees, 
near a regulated sector overall and at individual sites. 
This allows EJ analysis to move beyond looking at a small number of indicators 
for a few demographic groups, at one site in a single permitting decision, 
to a more complete picture of conditions near a whole set of facilities 
that is the focus of a risk analysis or a new regulation being considered, for example.

## EJAM features and benefits

EJAM provides a ready-to-use summary report, plus more flexibility, accuracy, and speed
than other tools have. The web-based app quickly provides results on the fly --
  The circular buffering module was optimized to be extremely fast
(allowing realtime exploratory work in an app),
while still using the block-population calculation EJScreen uses,
making it more consistent with EJScreen and more accurate than other approaches.

EJAM also lets one pick locations through several different approaches, such as 
- specifying facility points by industry categories of various types (NAICS, SIC, EPA program, etc.)
- providing a table of latitudes/longitudes of a list of points
- using shapefiles with polygons
- selecting Census units to compare, such as Counties, based on FIPS codes

The default demographic indicators are
[EJScreen basic demographic indicators](https://www.epa.gov/ejscreen/overview-demographic-indicators-ejscreen "https://www.epa.gov/ejscreen/overview-demographic-indicators-ejscreen"),
with the addition of race/ethnicity subgroups.

The data and software are shared as reusable,
well-documented functions or modules in an R package,
to allow software developers or analysts to take advantage of
these resources in running their own analyses or building or
supplementing their own tools, websites, or mobile apps.

## Updates

EJAM data updates are meant to match EJScreen scheduled updates and main version numbers,
so EJAM 2.2 uses the same basic data as EJScreen 2.2 (the version used from mid-2023
through early 2024). Version 2.2 had ACS 2017-2021 block group demographics
and Census 2020 block weights. The mid-2024 versions of EJScreen
and EJAM likewise use the same data, including ACS 2018-2022.

## Speed

The power of this tool enables faster and more accurate results than
other tools generally have been able to provide.
It also should be much more cost-effective and easy to use than the
standard geospatial analysis solutions would be.
This tool lets any user very quickly see an analysis of a very large number of places,
and immediately get a ready-to-use report that provides
perspective on an entire industrial sector or set of places.

## Accuracy

Compared to other often-used approaches,
EJAM high-resolution buffering provides more accurate information about which populations live inside a buffer,
which is important in rural areas where a single blockgroup can cover a very large area.
For circular buffers, EJAM uses the locations of internal points of Census 2020 blocks,
not areal apportionment of block groups, to estimate where residents live within each block group.
This avoids the simplistic assumption that people are evenly spread out within each block group.
Instead, it uses blocks to get information about which part of a block group is where residents actually live.
There are several million blocks in the US, as compared with fewer than a quarter million block groups.
The only more accurate approaches than what EJAM uses are
1) to use areal apportionment of blocks (not block groups), but that is very slow, or
2) to use the 30x30 meter grid EPA developed based on dasymetric estimates of where people live at even higher resolution than a block,
but that requires large amounts of storage and computer time.
EJAM was designed to replicate EJScreen methods and results for any given location,
such as the total population count within 1 mile, to maintain consistency and avoid any confusion.

EJAM also takes note of which residences are near which sites,
to avoid double-counting people in the summary statistics but still allow a user to view results for one site at a time.
This is something other tools and analyses often cannot provide -
  when they aggregate across sites they typically do not retain the statistics on individual sites,
and rarely if ever keep track of which communities are near multiple facilities.
Keeping track of this would also allow an analyst to explore how many people are near multiple sites,
or ask which sites in communities that already have multiple sites nearby.

EJAM was designed so that it can provide a continuous distribution of distances,
as distributed across blocks or people for one or all of the nearby facilities.
This enables exploration of the complete picture of proximities,
rather than using an arbitrary single distance defining near versus far.
The distribution can be sliced later for the summary statistics at any distance,
and can be summarized as a distribution of distances within each demographic group.
