---
title: Analysis of demographic and environmental conditions near selected facilities based on USEPA's EJAM tool
date: "`r Sys.Date()`"
params:
  authorname1: NA
  authoremail1: NA
  radius: NA
  where: NA
  sectorname_short: NA
  sitecount: NA
  acs_version: "2016-2020"
  ejscreen_version: "2.1"
  zonetype: NA
  in_the_x_zone: NA
  within_x_miles_of: NA
  facilities_studied: NA
  in_areas_where: NA
  risks_are_x: NA
  demog_how_elevated: NA
  envt_how_elevated: NA
  demog_high_at_what_share_of_sites: NA
  envt_high_at_what_share_of_sites: NA
  source_of_latlons: NA
  fundingsource: NA
  conclusion1: NA
  conclusion2: NA
  conclusion3: NA
author:
  - name: "`r params$authorname1`"
    email: "`r params$authoremail1`"
editor_options:
  markdown: wrap 72
linenumbers: false
keywords: 
  - environmental justice
  - socioeconomic
  - demographic
  - environmental pollutants
abstract: Abstract - Executive Order 14008 calls on EPA and other Agencies to make achieving environmental justice part of their missions, and EO 12989 directed EPA to make environmental justice part of their mission by identifying and addressing, as appropriate, disproportionately high and adverse human health or environmental effects of their programs, policies, and activities on minority populations and low-income populations in the United States. The United States Environmental Protection Agency (US EPA) analyzed baseline demographic and environmental conditions in communities living `r params$where` `r params$facilities_analyzed`. The analysis used EPA's EJAM tool and EJScreen version `r params$ejscreen_version` with demographic data based on the Census Bureau’s `r params$acs_version` American Community Survey (ACS). **The analysis found that**
output: 
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

Authors: `r params$authorname1`
Author contact info: `r params$authoremail1`

Radius: `r params$radius`

 
```{r, echo=FALSE}
in_the_places_analyzed <- {
    switch (
      params$zonetype,
    zone_is_named_x = params$in_the_x_zone, 
    zone_is_nearby = paste0(params$within_x_miles_of, params$facilities_studied),
    zone_is_risk_x = paste0(params$in_areas_where, params$risks_are_x)
    )}
```

# Executive Summary	 

Executive Order 14008 calls on EPA and other Agencies to make achieving
environmental justice part of their missions, and EO 12989 directed EPA
to make environmental justice part of their mission by identifying and
addressing, as appropriate, disproportionately high and adverse human
health or environmental effects of their programs, policies, and
activities on minority populations and low-income populations in the
United States.

EPA has conducted an analysis to characterize baseline environmental
conditions faced by communities living near `r params$sectorname_short`.
The United States Environmental Protection Agency (US EPA) analyzed
baseline demographic and environmental conditions in communities living
`r params$where`. The analysis used EPA's EJAM tool and EJScreen version
`r params$ejscreen_version` with demographic data based on the Census
Bureau's `r params$acs_version` American Community Survey (ACS).


## Broad overview of findings 

## Summary of Findings 

### Demographics overall 

### Demographics at key sites 

### Environment overall 

### Environment at key sites 

### Cumulative impacts at key sites 

# Introduction 

Executive Order 12898 (59 FR 7629; February 16, 1994) established
federal executive policy on environmental justice. Its main provision
directed federal agencies, to the greatest extent practicable and
permitted by law, to make environmental justice part of their mission by
identifying and addressing, as appropriate, disproportionately high and
adverse human health or environmental effects of their programs,
policies, and activities on minority populations and low-income
populations in the United States.

EPA defines environmental justice as the fair treatment and meaningful
involvement of all people regardless of race, color, national origin, or
income with respect to the development, implementation, and enforcement
of environmental laws, regulations, and policies.

Executive Order 14008 (86 FR 7619; January 27, 2021) also calls on
Agencies to make achieving environmental justice part of their missions
"by developing programs, policies, and activities to address the
disproportionately high and adverse human health, environmental,
climate-related and other cumulative impacts on disadvantaged
communities, as well as the accompanying economic challenges of such
impacts." It also declares a policy "to secure environmental justice and
spur economic opportunity for disadvantaged communities that have been
historically marginalized and overburdened by pollution and
under-investment in housing, transportation, water and wastewater
infrastructure and health care."

EPA also released its "Technical Guidance for Assessing Environmental
Justice in Regulatory Analysis" (U.S. EPA, 2016) to provide
recommendations that encourage analysts to conduct the highest quality
analysis feasible, recognizing that data limitations, time and resource
constraints, and analytic challenges will vary by media and
circumstance.


# Methods 

## Selection of sites analyzed 

## Estimating locations and population counts of residents 

### Spatial resolution of data 

### Analytic method for buffering, and tools used to implement that method 

## Demographic and environmental indicators 

# Findings 

## Text on Findings 

## Data Table 1 

## Data Table 2 

## Data Viz 1 – Barplot 

## Data Viz 2 – Map 

## Data Viz 3 – Boxplots 

# Notes on creating this document 

#### Table of EJAM Tool Parameter Values

```{r, warning=F, message=F, echo=F}
library(tidyverse)
param_table <- aggregate(values~ind, stack(params), toString) %>% 
  rename(Parameter = ind, 'Current Value' = values)

knitr::kable(param_table)
```

## Notes on formatting for a manuscript 

## Bibliography styles 

## Equations 

# Appendices 

## Competing interests 

## Author contributions 

## Acknowledgements 

## List of Abbreviations 

# References – from .bib file  

 

 