---
title: Analysis of demographic and environmental conditions near selected facilities
date: "`r Sys.Date()`"
params:
  testmode: TRUE
  analysis_title: "Analysis of demographic and environmental conditions near selected facilities"
  zonetype: NA
  in_the_x_zone: NA
  within_x_miles_of: NA
  facilities_studied: NA
  in_areas_where: NA
  risks_are_x: NA
  where: NA
  facilities_analyzed: "these sites"
  sectorname_short: NA
  sitecount: NA
  distance: NA
  total_pop: NA
  results: NA
  map: NA
  map_placeholder_png: "map_placeholder.png"
  envt_table: NA
  envt_table_placeholder_rda: "envt_table_placeholder.rda"
  envt_table_placeholder_png: "envt_table_placeholder.png"
  demog_table: NA
  demog_table_placeholder_rda:  "demog_table_placeholder.rda"
  demog_table_placeholder_png:  "demog_table_placeholder.png"
  boxplot: NA
  boxplot_placeholder_png: "boxplot_placeholder.png"
  barplot: NA
  barplot_placeholder_png: "barplot_placeholder.png"
  demog_how_elevated: NA
  envt_how_elevated: NA
  demog_high_at_what_share_of_sites: NA
  envt_high_at_what_share_of_sites: NA
  conclusion1: NA
  conclusion2: NA
  conclusion3: NA
  source_of_latlons: NA
  acs_version: "2016-2020"
  ejscreen_version: "2.1"
  fundingsource: NA
  authorname1: "USEPA"
  authoremail1: ""
author:
  - name: "`r params$authorname1`"
    email: "`r params$authoremail1`"
editor_options:
  markdown:
    wrap: 72
linenumbers: true
abstract: Abstract - Executive Order 14008 calls on EPA and other Agencies to make achieving environmental justice part of their missions, and EO 12989 directed EPA to make environmental justice part of their mission by identifying and addressing, as appropriate, disproportionately high and adverse human health or environmental effects of their programs, policies, and activities on minority populations and low-income populations in the United States. The United States Environmental Protection Agency (US EPA) analyzed baseline demographic and environmental conditions in communities living `r params$where` `r params$facilities_analyzed`. The analysis used EPA's EJAM tool and EJScreen version `r params$ejscreen_version` with demographic data based on the Census Bureau’s `r params$acs_version` American Community Survey (ACS). The analysis found that PLACEHOLDER EXAMPLE PLACEHOLDER EXAMPLE PLACEHOLDER EXAMPLE PLACEHOLDER EXAMPLE PLACEHOLDER EXAMPLE PLACEHOLDER EXAMPLE PLACEHOLDER EXAMPLE PLACEHOLDER EXAMPLE
output: 
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
bibliography: report_references.bib
---

```{r, echo=FALSE}

in_the_places_analyzed <- {
    switch (
      params$zonetype,
    zone_is_named_x = params$in_the_x_zone, 
    zone_is_nearby = paste0(params$within_x_miles_of, params$facilities_studied),
    zone_is_risk_x = paste0(params$in_areas_where, params$risks_are_x)
    )}
```


\newpage
# Executive Summary

Executive Order 14008 calls on EPA and other Agencies to make achieving
environmental justice part of their missions, and EO 12989 directed EPA
to make environmental justice part of their mission by identifying and
addressing, as appropriate, disproportionately high and adverse human
health or environmental effects of their programs, policies, and
activities on minority populations and low-income populations in the
United States.

EPA has conducted an analysis to characterize baseline environmental
conditions faced by communities living near `r params$sectorname_short`.
The United States Environmental Protection Agency (US EPA) analyzed
baseline demographic and environmental conditions in communities living
`r params$where`. The analysis used EPA's EJAM tool and EJScreen version
`r params$ejscreen_version` with demographic data based on the Census
Bureau's `r params$acs_version` American Community Survey (ACS).

## Broad overview of findings

PLACEHOLDER EXAMPLE The environmental indicators -- especially proximity
scores -- are more notable than the demographic indicators at these
sites. Demographic indicators are moderately above average overall.

PLACEHOLDER EXAMPLE The environmental indicators -- especially proximity
scores -- are more notable than the demographic indicators at these
sites. Demographic indicators are moderately above average overall.

PLACEHOLDER EXAMPLE The environmental indicators -- especially proximity
scores -- are more notable than the demographic indicators at these
sites. Demographic indicators are moderately above average overall.

## Summary of Findings

-   PLACEHOLDER EXAMPLE Overall, % people of color, % limited English
    proficiency, and the Demographic Indicator are more than 1.5x the
    State rate, for the population within 1 mile.

-   PLACEHOLDER EXAMPLE About a third of these sites are above the 80th
    percentile in State for the Demographic Indicator. The same is true
    for % low income and % with less than high school.

-   PLACEHOLDER EXAMPLE About half of these 57 sites are in just 4
    states: FL, NY, PA, or MA. Most of the people here live near just 8
    sites (15% of sites). Most of the sites with the higher demographic
    indicators are owned by Covanta or Wheelabrator.

-   PLACEHOLDER EXAMPLE Many of the sites with the highest demographic
    indicators also have proximity scores that are 5 to 10 times the
    State average.

    -   PLACEHOLDER EXAMPLE The average person's RMP score is more than
        3x their State's average.
    -   PLACEHOLDER EXAMPLE The average person's NPL and TSDF scores are
        about 2.5x State averages.
    -   PLACEHOLDER EXAMPLE Most of these sites are \>=80th in State for
        NATA. Same for RMP.
 

- PLACEHOLDER EXAMPLE

- PLACEHOLDER EXAMPLE

- PLACEHOLDER EXAMPLE 




# Introduction

Executive Order 12898 (59 FR 7629; February 16, 1994) established
federal executive policy on environmental justice. Its main provision
directed federal agencies, to the greatest extent practicable and
permitted by law, to make environmental justice part of their mission by
identifying and addressing, as appropriate, disproportionately high and
adverse human health or environmental effects of their programs,
policies, and activities on minority populations and low-income
populations in the United States.

EPA defines environmental justice as the fair treatment and meaningful
involvement of all people regardless of race, color, national origin, or
income with respect to the development, implementation, and enforcement
of environmental laws, regulations, and policies.

Executive Order 14008 (86 FR 7619; January 27, 2021) also calls on
Agencies to make achieving environmental justice part of their missions
"by developing programs, policies, and activities to address the
disproportionately high and adverse human health, environmental,
climate-related and other cumulative impacts on disadvantaged
communities, as well as the accompanying economic challenges of such
impacts." It also declares a policy "to secure environmental justice and
spur economic opportunity for disadvantaged communities that have been
historically marginalized and overburdened by pollution and
under-investment in housing, transportation, water and wastewater
infrastructure and health care."

EPA also released its "Technical Guidance for Assessing Environmental
Justice in Regulatory Analysis" (U.S. EPA, 2016) to provide
recommendations that encourage analysts to conduct the highest quality
analysis feasible, recognizing that data limitations, time and resource
constraints, and analytic challenges will vary by media and
circumstance.

# Methods

EPA's EJAM (Environmental Justice Analysis Multisite) tool was used to
develop this analysis. EPA's [EJAM tool](https://rstudio-connect.dmap-stage.aws.epa.gov/content/dc3cda00-20a2-47ed-a753-0dcb89eb8f2a/ "Environmental Justice Analysis Multisite (EJAM) tool") is a user-friendly web app that can
summarize demographics and environmental conditions for any list of
places in the nation. It provides interactive results and a formatted,
ready-to-share report with written explanations of the results, tables,
and graphics. The report can provide EJ-related information about people
who live in communities near any of the industrial facilities on a list,
for example.

The basic methodology and data used for this analysis are the same as
those used in EPA's EJScreen tool, with a few exceptions described
below, and are described in detail in EJScreen's documentation, at
[EJScreen](https://epa.gov/ejscreen "EPA EJScreen homepage") and with
more technical details available at [EJScreen technical documentation
page](https://www.epa.gov/ejscreen/technical-information-about-ejscreen "EPA EJScreen technical information").
The only notable differences are the following:

- EJAM may include additional demographic or environmental indicators.
- For a proximity analysis (to characterize everyone living within a certain certain distance from a point such as a facility), EJAM identifies which residents live nearby using a slight variation on how the distance to each Census block is measured. While EJScreen uses ESRI's ArcGIS calculations, EJAM calculates the distance using formulas implemented in the R language for statistical computing [@R-base]. These measurements provide almost identical results for estimated distance from the average person in a block to a given site point. PLACEHOLDER MORE INFO
- EJAM aggregates indicator values within and across locations, converts them to percentiles, and does other summary calculations using the same formulas to the greatest extent possible, but in R rather than within EJScreen itself. There may be slight differences between raw scores and percentiles in EJScreen and EJAM in some cases.
 
## Selection of sites analyzed


## Estimating locations and population counts of residents

### Spatial resolution of data

The analysis used EPA's EJAM tool and EJScreen version `r params$ejscreen_version` with demographic data based on the Census Bureau's `r params$acs_version` American Community Survey (ACS) and the corresponding version of Decennial Census information on geographic boundaries and FIPS codes for blocks, block groups, tracts, counties, and States.

See EJScreen methodology for details. 

### Analytic method for buffering, and tools used to implement that method

See EJScreen methodology for details. EJAM finds nearby block points using a very fast approach but otherwise uses the EJScreen method of estimating nearby residential populations for proximity analysis.

## Demographic and environmental indicators

PLACEHOLDER - The demographics included here are those in EJScreen and also race/ethnicity subgroups that comprise the total count of people of color. POC are defined as all other than those self-identifying in ACS survey data as white, single race, not Hispanic or Latino - i.e., non-hispanic white alone (NHWA). The subgroups include Hispanic or Latino ("hispanic"), several groups that are not hispanic but of only single race (e.g., Asian, or more specifically non-hispanic asian alone), non-hispanic other single race, and non-hispanic multiracial. 

See EJScreen methodology for details. 

# Findings


## Text on Findings


### Demographics overall

Almost all the EJ-relevant groups (low-income, people of color, etc.)
are at least somewhat over-represented near these sites overall (at the
collection of sites as a whole).

Most notably, % with limited English, % low income, and % with less than
high school education near these sites are about 1.5 to 1.7 times the US
overall rates. 

The people living near these sites are 40% more likely to be in Limited-English Households than the average US resident.

The % with limited English is driven by high scores at
only a few very highly population sites -- it is high enough that the
rate overall is in the top quintile nationally (83 percentile), but
other demographics do not reach the top quintile for the entire
population across all sites as a whole.

Near most of these 72 sites, % low income is at least 1.3x the rate in
the US overall, and near 1 in 4 it is at least 1.5x the US rate.

### Demographics at key sites

There are a handful of sites each with at least one very high
demographic stat within 1 mile, however this may be within the normal
range of what one would expect across the range of US residential areas
\-- there does not appear to be a pattern of an unusually large share of
these 72 sites having any given demographic stat in the top 5%, for
example.

Seven key sites have at least some demographic percentages more than 2x
the US average: Crawford in Chicago IL (densely populated location), Bay
Shore in Ohio, Watts Bar Fossil Plant in Spring City TN (but has almost
no nearby residents), Arkwright in Macon GA, Venice in IL, Lake Shore in
Cleveland, and Fair Station in Muscatine IA.

At two sites, percent people of color and percent low income are both
more than twice the US averages (Venice and Lake Shore).

A few sites have over triple the US average unemployment rate (two of
which are in the top 5% nationwide for their rate of unemployment, Bay
Shore and Watts Bar).

### Environment overall

The people living nearby these sites as a whole are facing relatively
high exposure to indicators of RMP proximity and possible lead paint due
to older buildings. Overall the average person nearby has RMP proximity
more than 3 times the US average. Lead paint and traffic are also
notable, at more than 2x the US average. The average person near any of
these sites lives in a blockgroup that is at the 80th percentile (worst
quintile) of the US for RMP and lead paint -- that is unusual because it
is a pattern for these populations as a whole not just at one site. The
same is almost true for traffic and UST -- the average person nearby has
those indicators in the worst quartile of the US. (Wastewater and
Superfund also tend to be very high at an unusually large share of these
sites but not necessarily at the ones in highly populated areas).

### Environment at key sites

There are an unusually large number of sites with very high
environmental stats within 1 mile. Surprisingly, 27 of these 72 sites
have at least one above the 95th percentile. One might expect 5% of
these sites (i.e., 3.6 sites) to have a given score in the top 5%
nationwide, but there are 8 sites (2x what one might expect) with RMP
proximity scores in the top 5%, and the same is true for Superfund NPL
proximity scores (there are 8 sites \>=95th %ile instead of the expected
3.6). For the wastewater discharge indicator, there are 11 such sites,
over 3x as many as one might expect. Most sites have at least one
environmental indicator \>80th percentile.

### Cumulative impacts at key sites

Multiple environmental stressors are also an issue in some cases - At
two of the sites, there are five environmental indicators that are all
more than twice the US average (Valley and Crawford).


discussion of map 

discussion of map 

discussion of map 

discussion of map 

discussion of map 


## Data Viz 1 -- Map


```{r map, echo=FALSE,  out.height='500px', out.width='800px'}
if ("NA" == (params$map)) {
   (knitr::include_graphics(path = params$map_placeholder_png))
} else {
  print( params$map )
}
```



discussion of plot goes here


discussion of plot goes here


discussion of plot goes here


discussion of plot goes here



## Data Viz 2-- Barplot


```{r barplot, echo=FALSE, fig.align='center', out.height='500px', out.width='800px'}
if ("NA" == (params$barplot)) {
 knitr::include_graphics(path = params$barplot_placeholder_png)
} else {
  params$barplot  
}
```

 
discussion of table here 


discussion of table here 



discussion of table here 


discussion of table here 


## Data Table 1. Demographic Indicators

```{r demog_table, echo=FALSE, out.width='900px'}
if ("NA" != (params$demog_table)) {
params$demog_table
} else {
  knitr::include_graphics(path = params$demog_table_placeholder_png)
  # not sure how to do this here:
  #  print(gt::render_gt(params$demog_table_placeholder_rda))
}
```


discussion of table here 


discussion of table here 


discussion of table here 


discussion of table here 


discussion of table here 



## Data Table 2. Environmental Indicators

```{r envt_table, echo=FALSE, out.width='900px'}
if ("NA" !=  (params$envt_table)) {
params$envt_table
} else {
  knitr::include_graphics(path = params$envt_table_placeholder_png)
  # not sure how to do this here:
  #  print(gt::render_gt(params$envt_table_placeholder_rda))
}
```



discussion of plot here 


discussion of plot here 


discussion of plot here 


discussion of plot here 


discussion of plot here 



## Data Viz 3 -- Boxplots

```{r boxplot, echo=FALSE, fig.align='center', out.height='500px', out.width='800px'}
if ("NA" == (params$boxplot)) {
 knitr::include_graphics(path = params$boxplot_placeholder_png)
} else {
  params$boxplot +
  ggplot2::scale_x_discrete(labels = function(x) stringr::str_wrap(x, 10)) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust=0.5, size=12),
        plot.subtitle = ggplot2::element_text(hjust=0.5, size=11),
        axis.title.y = ggplot2::element_text(size=10),
        axis.text = ggplot2::element_text(size=9)
        )
}
```

\newpage
# Appendices


## Acknowledgements


## List of Abbreviations


\newpage
# References

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE }

# <!-- -- References can come from report_bibliography.bib file -->
# 
# <!-- see citation()  and in Visual mode of report.Rmd editor in RStudio, see Insert Citation -->
# <!-- Depends:  -->
# <!--     R (>= 2.10),  -->
# <!--     shiny (>= 1.7.2), -->
# <!--     EJAMblockdata, -->
# <!--     EJAMfrsdata, -->
# <!--     EJAMbatch.summarizer, -->
# <!--     EJAMejscreenapi -->
# <!-- Imports: -->
# <!--   attempt, -->
# <!--   collapse, -->
# <!--   config (>= 0.3.1), -->
# <!--   DBI, -->
# <!--   doSNOW, -->
# <!--   DT, -->
# <!--   EJAMejscreendata, -->
# <!--   foreach, -->
# <!--   ggplot2, -->
# <!--   glue, -->
# <!--   golem (>= 0.3.3), -->
# <!--   htmltools, -->
# <!--   leaflet, -->
# <!--   magrittr, -->
# <!--   openxlsx, -->
# <!--   pdist, -->
# <!--   pkgload, -->
# <!--   readxl, -->
# <!--   rmarkdown, -->
# <!--   RMySQL, -->
# <!--   SearchTrees, -->
# <!--   shinyBS, -->
# <!--   shinycssloaders, -->
# <!--   shinyjs, -->
# <!--   sf, -->
# <!--   sp, -->
# <!--   tidyverse, -->
# <!--   data.table -->
# <!-- Remotes:  -->
# <!--     github::USEPA/EJAMblockdata, -->
# <!--     github::USEPA/EJAMfrsdata, -->
# <!--     github::USEPA/EJAMejscreenapi, -->
# <!--     github::USEPA/EJAMbatch.summarizer, -->
# <!--     github::USEPA/EJAMejscreendata -->
# <!-- Suggests:  -->
# <!--     knitr, -->
# <!--     spelling, -->
# <!--     testthat (>= 3.0.0) -->

```




```{r eval=params$testmode, message=FALSE, warning=FALSE, include=FALSE}
# Developer notes on creating this document

#### Developer notes on EJAM Tool Parameter Values
library(tidyverse)
ind=NA;values=NA # (just to prevent RStudio warning these are not in scope)
param_table <- aggregate(values~ind, stack(params), toString) %>% 
  dplyr::rename(Parameter = ind, 'Current Value' = values)
#knitr::kable(param_table)
param_table
```

