---
title: Analysis of demographic and environmental conditions near selected facilities based on USEPA's EJAM tool
date: "`r Sys.Date()`"
params:
  analysis_title: NA
  authorname1: NA
  authoremail1: NA
  where: NA
  sectorname_short: NA
  sitecount: NA
  distance: NA
  total_pop: NA
  results: NA
  map: NA
  envt_table: NA
  demog_table: NA
  boxplot: NA
  zonetype: NA
  in_the_x_zone: NA
  within_x_miles_of: NA
  facilities_studied: NA
  in_areas_where: NA
  risks_are_x: NA
  demog_how_elevated: NA
  envt_how_elevated: NA
  demog_high_at_what_share_of_sites: NA
  envt_high_at_what_share_of_sites: NA
  conclusion1: NA
  conclusion2: NA
  conclusion3: NA
  source_of_latlons: NA
  acs_version: "2016-2020"
  ejscreen_version: "2.1"
  fundingsource: NA
author:
  - name: "`r params$authorname1`"
    email: "`r params$authoremail1`"
editor_options:
  markdown: wrap 72
    wrap: 72
linenumbers: false
keywords:
  - environmental justice
  - socioeconomic
  - demographic
  - environmental pollutants
abstract: Abstract - Executive Order 14008 calls on EPA and other Agencies to make achieving environmental justice part of their missions, and EO 12989 directed EPA to make environmental justice part of their mission by identifying and addressing, as appropriate, disproportionately high and adverse human health or environmental effects of their programs, policies, and activities on minority populations and low-income populations in the United States. The United States Environmental Protection Agency (US EPA) analyzed baseline demographic and environmental conditions in communities living `r params$where` `r params$facilities_analyzed`. The analysis used EPA's EJAM tool and EJScreen version `r params$ejscreen_version` with demographic data based on the Census Bureau’s `r params$acs_version` American Community Survey (ACS). **The analysis found that...................**
output: 
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

Authors: `r params$authorname1` Author contact info: `r params$authoremail1`

Radius: `r params$distance`

```{r, echo=FALSE}
in_the_places_analyzed <- {
    switch (
      params$zonetype,
    zone_is_named_x = params$in_the_x_zone, 
    zone_is_nearby = paste0(params$within_x_miles_of, params$facilities_studied),
    zone_is_risk_x = paste0(params$in_areas_where, params$risks_are_x)
    )}
```

# Executive Summary

Executive Order 14008 calls on EPA and other Agencies to make achieving environmental justice part of their missions, and EO 12989 directed EPA to make environmental justice part of their mission by identifying and addressing, as appropriate, disproportionately high and adverse human health or environmental effects of their programs, policies, and activities on minority populations and low-income populations in the United States.

EPA has conducted an analysis to characterize baseline environmental conditions faced by communities living near `r params$sectorname_short`. The United States Environmental Protection Agency (US EPA) analyzed baseline demographic and environmental conditions in communities living `r params$where`. The analysis used EPA's EJAM tool and EJScreen version `r params$ejscreen_version` with demographic data based on the Census Bureau's `r params$acs_version` American Community Survey (ACS).

## Broad overview of findings

The environmental indicators -- especially proximity scores -- are more notable than the demographic indicators at these sites. Demographic indicators are moderately above average overall.

## Summary of Findings

-   Overall, % people of color, % limited English proficiency, and the Demographic Indicator are more than 1.5x the State rate, for the population within 1 mile.

-   About a third of these sites are above the 80th percentile in State for the Demographic Indicator. The same is true for % low income and % with less than high school.

-   About half of these 57 sites are in just 4 states: FL, NY, PA, or MA. Most of the people here live near just 8 sites (15% of sites). Most of the sites with the higher demographic indicators are owned by Covanta or Wheelabrator.

-   Many of the sites with the highest demographic indicators also have proximity scores that are 5 to 10 times the State average.

    -   The average person's RMP score is more than 3x their State's average.
    -   The average person's NPL and TSDF scores are about 2.5x State averages.
    -   Most of these sites are \>=80th in State for NATA. Same for RMP.

### Demographics overall

### Demographics at key sites

### Environment overall

### Environment at key sites

### Cumulative impacts at key sites

# Introduction

Executive Order 12898 (59 FR 7629; February 16, 1994) established federal executive policy on environmental justice. Its main provision directed federal agencies, to the greatest extent practicable and permitted by law, to make environmental justice part of their mission by identifying and addressing, as appropriate, disproportionately high and adverse human health or environmental effects of their programs, policies, and activities on minority populations and low-income populations in the United States.

EPA defines environmental justice as the fair treatment and meaningful involvement of all people regardless of race, color, national origin, or income with respect to the development, implementation, and enforcement of environmental laws, regulations, and policies.

Executive Order 14008 (86 FR 7619; January 27, 2021) also calls on Agencies to make achieving environmental justice part of their missions "by developing programs, policies, and activities to address the disproportionately high and adverse human health, environmental, climate-related and other cumulative impacts on disadvantaged communities, as well as the accompanying economic challenges of such impacts." It also declares a policy "to secure environmental justice and spur economic opportunity for disadvantaged communities that have been historically marginalized and overburdened by pollution and under-investment in housing, transportation, water and wastewater infrastructure and health care."

EPA also released its "Technical Guidance for Assessing Environmental Justice in Regulatory Analysis" (U.S. EPA, 2016) to provide recommendations that encourage analysts to conduct the highest quality analysis feasible, recognizing that data limitations, time and resource constraints, and analytic challenges will vary by media and circumstance.

# Methods

## Selection of sites analyzed

## Estimating locations and population counts of residents

### Spatial resolution of data

### Analytic method for buffering, and tools used to implement that method

## Demographic and environmental indicators

# Findings

## Text on Findings

## Data Table 1

```{r, out.width ='900px'}
 # dt <- cbind('Value at Uploaded Sites' = as.list( 100*round(params$results$results_overall[ , ..names_d], 2)),
 #          'State Percentile' = as.list(round(params$results$results_overall[, ..names_d_state_pctile], 2)),
 #          'National Percentile' = as.list(round(params$results$results_overall[, ..names_d_pctile], 2)))
 #    
 #  rownames(dt) <- EJAMbatch.summarizer::names_d_friendly[pmatch(names_d, EJAMbatch.summarizer::names_d_batch)]
 #  rownames(dt)[1] <- 'Demographic Index'
 #  
 #  dt <- as.data.frame(dt) %>% rownames_to_column(var = 'Indicator')
 #    
 #  DT::datatable(dt, rownames = FALSE, 
 #                options = list(autoWidth = TRUE))
params$demog_table
```

## Data Table 2

## Data Viz 1 -- Barplot

## Data Viz 2 -- Map

## Data Viz 3 -- Boxplots

```{r boxplot, out.width = '800px', out.height = '500px', fig.align='center'}
params$boxplot +
  ggplot2::scale_x_discrete(labels = function(x) stringr::str_wrap(x, 10)) +
  theme(plot.title = element_text(hjust=0.5, size=12),
        plot.subtitle = element_text(hjust=0.5, size=11),
        axis.title.y = element_text(size=10),
        axis.text = element_text(size=9)
        )
```

# Notes on creating this document

#### Table of EJAM Tool Parameter Values

```{r, warning=F, message=F, echo=F}
library(tidyverse)
param_table <- aggregate(values~ind, stack(params), toString) %>% 
  rename(Parameter = ind, 'Current Value' = values)

#knitr::kable(param_table)
param_table
```

## Notes on formatting for a manuscript

## Bibliography styles

## Equations

# Appendices

## Competing interests

## Author contributions

## Acknowledgements

## List of Abbreviations

# References -- from .bib file
